The Small World Network Effect in Software Project Teams \author{Kevin Peterson\\ \small \texttt{pete1968@umn.edu} }

### 2013-12-05 paper/interview1.tex
Proofread.

\subsection*{Q3: When assembling a team, which is preferred -- members familiar with each other, or members familiar with the task or project.}
Familiarity is good, but you have to account for burnout if people work on the same project for too long. This really depends on project dynamics, special skills needed, etc. It is important to note, however, that team members can be familiar with each other but not get along well, or not work well with each other. I have had projects where team members were very familiar with each other, but did not function well as a team. Clashing personalities and other team problems can cause a toxic environment. Most importantly, whether the team is familiar with each other or not, is that they can share knowledge and skills.

>>>>

\subsection*{Q3: When assembling a team, which is preferred -- members familiar with each other, or members familiar with the task or project?}
Familiarity is good, but you have to account for burnout if people work on the same project for too long. This really depends on project dynamics, special skills needed, etc. It is important to note, however, that team members can be familiar with each other but not get along well, or not work well with each other. I have had projects where team members were very familiar with each other, but did not function well as a team. Clashing personalities and other team problems can cause a toxic environment. Most importantly, whether the team is familiar with each other or not, is that they can share knowledge and skills.

### 2013-12-05 paper/interview2.tex
Proofread.

\subsection*{Q3: When assembling a team, which is preferred -- members familiar with each other, or members familiar with the task or project.}
This really depends on the time-line of the project. Short-term projects don't have as much time for team bonding -- it is more important to get the people with the right skill set. For longer term projects, you have to be much more careful about people getting along with each other. for these types of projects, you can afford to invest some time into getting the team right and doing some team building. Also, with long term projects, you have more of a chance of members ``learning on the job," so even if the technical skills aren't there right away, you can factor in some learning time if they are a good fit personality-wise. It is the same reason why we ask a lot of behavioral questions for new full-time employees, and less so with contractors. It all depends on the time commitment and how much team building you can plan for.

>>>>

\subsection*{Q3: When assembling a team, which is preferred -- members familiar with each other, or members familiar with the task or project?}
This really depends on the time-line of the project. Short-term projects don't have as much time for team bonding -- it is more important to get the people with the right skill set. For longer term projects, you have to be much more careful about people getting along with each other. for these types of projects, you can afford to invest some time into getting the team right and doing some team building. Also, with long term projects, you have more of a chance of members ``learning on the job," so even if the technical skills aren't there right away, you can factor in some learning time if they are a good fit personality-wise. It is the same reason why we ask a lot of behavioral questions for new full-time employees, and less so with contractors. It all depends on the time commitment and how much team building you can plan for.

### 2013-12-05 paper/main.tex
Proofread.

\begin{abstract}
Team cohesion and the dynamics of team forming are important parts of any project, with software projects being no exception. An interesting aspect of team building is the relationships formed between the team members. Because of these relationships, visualizing software team members as a graph may be a natural way to explore these relationships. As team members move between projects, these graphs become more and more connected as team members collaborate and form relationships. We show that this connectivity, known as the ``small world effect," has a positive impact on team performance when the connectivity levels are moderate. Performance degrades, however, at both very high and very low levels of connectivity. This aligns with similar research findings of non-software teams.
\end{abstract}

@@

\section{Introduction}
A social network is a graph of people and the connections between them. The dynamics of social networks have proved to be an important research topic for many different areas of study, from academic publication\cite{barabasi2002evolution}, to the success of Broadway musicals\cite{uzzi2005collaboration}. In general, a social network is important to understanding how ideas and influence are spread\cite{kempe2003maximizing}. Given that, we can begin to investigate the best conditions for optimal group performance within these networks.

A ``small world'' network is a social network characterized by high clustering of graph nodes paired with a short average path between nodes\cite{watts1998collective}. ``Clustering'' in this context means how closely nodes in the graph are related to each other, or ``the friend of my friend is also my friend.'' ``Average path length'' is the the shortest route between all possible nodes, averaged over the entire graph. Milgram, in his seminal study, found that any two people are linked through a chain of friends and acquaintances on average of 6 people long\cite{milgram1967small}, or in other words, the average path length between any two people.

@@

\subsection{Collection and Storage}
Data was collected using the freely available FLOSSmole\cite{floss2006} data sets of free and open source projects. Two datasets were analyzed -- data from from the popular open source project site \textit{Freecode}\footnote{http://freecode.com/}, and another from open source software repository \textit{SourceForge}\footnote{http://sourceforge.net/}. The Freecode dataset contained data up to September, 2013, while the SourceForge dataset was slightly older with data gathered on June 2009. Both datasets were chosen because they met two basic criteria: (1) they contained an exhaustive list of projects and contributors to those projects, and (2) they provided some sort of quantitative measure of project ``popularity." This popularity metric will be further described in sections below. Results were analyzed using the Python graph processing packaged NetworkX\cite{hagberg-2008-exploring}. All source code pertaining to the collection and analysis can be found at \url{https://github.com/kevinpeterson/small-world-effect-research}.

\subsection{Analysis}
The collected data was organized into a graph structure, with the nodes representing the contributors and the edges representing one or more shared project between contributors. For our purposes, ``contributors'' includes any person that has been identified as being associated with the project by the project hosting site (Freecode and SourceForge, for our purposes). It is important to note that ``contributor'' does not always imply code contributions -- as issue reporting and documentation, among other things, are valid contributions. Various properties of the graph were analyzed, followed by a qualitative section to help link the quantitative analysis to subject matter expert experiences.

\subsubsection{Graph Structure}
For each dataset, we first considered the dataset to be one large graph. It was observed that the contributor graph was ``disconnected,'' meaning that there was not a path between all possible nodes. From this large disconnected graph, we can extract a set of connected subgraphs, where each connected subgraph was analyzed. Metrics were then computed for each of these subgraphs. In order to correctly compute the metrics below, a minimum subgraph size was required. Specifically, many of the graph metrics of interest measured ``triangles'' of nodes and their connections. This implies a subgraph of {$>= 3$} nodes is needed. Subgraphs of {$< 3$} nodes were not considered for analysis.

@@

The contributor/project data can be represented as a bipartite graph. A bipartite graph is a specialized type of graph containing two disjoint sets of verticies. This can be denoted by ${G=(P,C,E)}$, where ${P \cap C = \emptyset}$. For our purposes, let set $P$ be the set of all projects, and set $C$ be the set of all contributors. Each project node, therefore, is only connected to contributor nodes, and vice versa, as show in figure \ref{fig:example_bipartite_graph}. This bipartite representation is typical of what is found in social and collaboration networks\cite{ramasco2004self}.

A bipartite graph projection is necessary to further analyze the graph. A bipartite projection involves taking the disjoint node sets $P$ and $C$, and representing the graph as relationships between only one of those sets of nodes. A projection onto the contributor nodes, or $C$, of figure \ref{fig:example_bipartite_graph} is show in figure \ref{fig:example_bipartite_projection_graph}. Here, contributors are connected directly, with edges occurring if a two contributors share one or more common projects. It is important to note that the projection could have been done in terms of the projects, not the contributors. This would have led to a graph with nodes of projects $P$, each being linked by sharing a common contributor. This approach is a common way of representing these bipartite graphs\cite{newman2001scientific}, although it is known that bipartite projections are lossy compared to the data represented in the original graph\cite{zhou2007bipartite}.\\

@@

\noindent\\\textit{\textbf{Clustering Coefficient - Transitivity ($C^\Delta$)}}\\
We calculate the \textit{Clustering Coefficient} (or $C^\Delta$) as a measure of the proportion of closed triangles in a graph\cite{newman2003structure}:
\[C^\Delta = 3 \times \frac{\text{number of triangles}}
                    {\text{number of connected triples}}\]

This metric shows us how closely related, or ``clustered,'' that the graph is. It is also the probability that two nodes will be connected if they share a common neighbor\cite{newman2003properties}. In our context, if contributors $C_1$ and $C_2$ both collaborate with a common contributor w $C_3$, this metric represents the probability that $C_1$ and $C_2$ will collaborate.

@@

From this \textit{Local Clustering Coefficient} $C_i$, we can then take the average over the entire set of nodes in the graph.
\[ \overline{C^\lambda} = \frac{1}{|P|}\sum^{|P|}_{i=1}C^{\lambda}_i \]

@@

Where {$v_i \in P$}, {$v_j \in P$}, and {$\lambda(v_i,v_j)$} denotes the shortest path between the two nodes. We only considered path length for connected subgraphs, avoiding some complexities of this calculation on disconnected graphs\cite{boccaletti2006complex}.

@@

We assume for throughout that any graph exhibiting {$Q > 1$} is a small world graph.

@@

To further study these factors, we arranged interviews with two current professionals in the field. Both have had experience with project management and are certified Project Management Professionals (PMP)\textregistered. We proposed to them the following questions, with the intent of expanding further the questions posed above:

\noindent\\\textit{Q1: How do development teams collaborate and share ideas?}\\
\noindent\\\textit{Q2: Do teams generally remain intact over multiple projects, or do members often shuffle and re-combine?}\\
\noindent\\\textit{Q3: When assembling a team, which is preferred -- members familiar with each other, or members familiar with the task or project.}\\
\noindent\\\textit{Q4: What are some characteristics of a high performing team?}\\
\noindent\\\textit{Q5: In your experience, is it better to (A) keep an innovative team intact, (B) periodically bring in new members or re-assign existing members, or (C) split up and disperse the team to as many other projects as possible?}\\

@@

Summary statistics for the graph analysis are captured in figure \ref{fig:summary_stats}.  Statistics analyzed included:\\
$Q$ - \textit{Small World Quotient}\\
$C^\Delta$ - \textit{Clustering Coefficient - Transitivity}\\
$\overline{C^\lambda}$ - \textit{Clustering Coefficient - Average}\\
$L$ - \textit{Average Path Length}\\
$N$ - \textit{Number of Nodes}\\
$E$ - \textit{Number of Edges}\\
The results shown were as measured over the number of projects $P$, with the total number of contributors $C$. The number of connected subgraphs analyzed is shown as $P_c$. The connected subgraphs are the main unit of analysis, and each statistical operation is applied to the set of subgraphs $P_c$.

@@

Distribution of $Q$ values are shown in figures \ref{fig:q_fc_distribution} and \ref{fig:q_sf_distribution}. The values of $Q$ generally follow a power-law distribution, with most values being at our around {$Q = 1$}. This indicates that many graphs are close to the delineation of being considered small-world\cite{humphries2008network}.

@@

\subsection{Qualitative}
The research interviews provided a look into how some of these quantitative findings translate into actual industry practice. They also helped us to obtain a deeper understanding of what we had found in the data analysis, as we could begin to build some workplace context into our results. There were are few re-occurring themes from which we will base our analysis\cite{stmartin_interview,rana_interview}:

@@

It was also indicated that priority was a major factor. Different project priorities, and the resultant shifting of resources, had the possible effect of the weakening of string team's core members. This can be related to focus as well, as changing priorities can cause resource realignment and decreased overall focus.

>>>>

\begin{abstract}
Team cohesion and the dynamics of team formation are important parts of any project, with software projects being no exception. An interesting aspect of team building is the relationships formed between the team members. Because of these relationships, visualizing software team members as a graph may be a natural way to explore these relationships. As team members move between projects, these graphs become more and more connected as team members collaborate and form relationships. We show that this connectivity, known as the ``small world effect," has a positive impact on team performance when the connectivity levels are moderate. Performance degrades, however, at both very high and very low levels of connectivity. This aligns with similar research findings of non-software teams.
\end{abstract}

@@

\section{Introduction}
A social network is a graph of people and the connections between them. The dynamics of social networks have proven to be an important research topic for many different areas of study, from academic publication\cite{barabasi2002evolution}, to the success of Broadway musicals\cite{uzzi2005collaboration}. In general, a social network is important to understanding how ideas and influence are spread\cite{kempe2003maximizing}. Given that, we can begin to investigate the best conditions for optimal group performance within these networks.

A ``small world'' network is a social network characterized by high clustering of graph nodes paired with a short average path between them\cite{watts1998collective}. ``Clustering'' in this context means how closely nodes in the graph are related to each other, or ``the friend of my friend is also my friend.'' ``Average path length'' is the shortest route between all possible nodes, averaged over the entire graph. Milgram, in his seminal study, found that any two people are linked through a chain of friends and acquaintances on average of 6 people long\cite{milgram1967small}, or in other words, the average path length between any two people is 6.

@@

\subsection{Collection and Storage}
Data was collected using the freely available FLOSSmole\cite{floss2006} datasets of free and open source projects. Two datasets were analyzed -- data from from the popular open source project site \textit{Freecode}\footnote{http://freecode.com/}, and another from open source software repository \textit{SourceForge}\footnote{http://sourceforge.net/}. The Freecode dataset contained data up to September, 2013, while the SourceForge dataset was slightly older with data gathered on June 2009. Both datasets were chosen because they met two basic criteria: (1) they contained an exhaustive list of projects and contributors to those projects, and (2) they provided some sort of quantitative measure of project ``popularity." This popularity metric will be further described in sections below. Results were analyzed using the Python graph processing packaged NetworkX\cite{hagberg-2008-exploring}. All source code pertaining to the collection and analysis can be found at \url{https://github.com/kevinpeterson/small-world-effect-research}.

\subsection{Analysis}
The collected data was organized into a graph structure, with the nodes representing the contributors and the edges representing one or more shared project between contributors. For our purposes, ``contributors'' includes any person that has been identified as being associated with the project by the project hosting site (Freecode and SourceForge). It is important to note that ``contributor'' does not always imply code contributions -- as issue reporting and documentation, among other things, are valid contributions. Various properties of the graph were analyzed, followed by a qualitative section to help link the quantitative analysis to subject matter expert experiences.

\subsubsection{Graph Structure}
For each dataset, we first considered the dataset to be one large graph. It was observed that the contributor graph was ``disconnected,'' meaning that there was not a path between all possible nodes. From this large disconnected graph, we extracted a set of connected subgraphs, where each connected subgraph was analyzed. Metrics were then computed for each of these subgraphs. In order to correctly compute the metrics below, a minimum subgraph size was required. Specifically, many of the graph metrics of interest measured ``triangles'' of nodes and their connections. This implies a subgraph of {$>= 3$} nodes is needed. Subgraphs of {$< 3$} nodes were not considered for analysis.

@@

The contributor/project data can be represented as a bipartite graph. A bipartite graph is a specialized type of graph containing two disjoint sets of verticies. This can be denoted by ${G=(P,C,E)}$, where ${P \cap C = \emptyset}$. For our purposes, let set $P$ be the set of all projects, set $C$ be the set of all contributors, and $E$ the edges that connect them. Each project node, therefore, is only connected to contributor nodes, and vice versa, as show in figure \ref{fig:example_bipartite_graph}. In other words, the sets $P$ and $C$ are disjoint. This bipartite representation is typical of what is found in social and collaboration networks\cite{ramasco2004self}.

A bipartite graph projection is necessary to further analyze the graph. A bipartite projection involves taking the disjoint node sets $P$ and $C$, and representing the graph as relationships between only one of those sets of nodes. A projection onto the contributor nodes, or $C$, of figure \ref{fig:example_bipartite_graph} is show in figure \ref{fig:example_bipartite_projection_graph}. Here, contributors are connected directly, with edges occurring if a two contributors share one or more common projects. It is important to note that the projection could have been done in terms of the projects, instead of the contributors. This would have led to a graph with nodes of projects $P$, each being linked by sharing a common contributor. Projection is a common way of representing these bipartite graphs\cite{newman2001scientific}, although it is known that bipartite projections are lossy compared to the data represented in the original graph\cite{zhou2007bipartite}.\\

@@

\noindent\\\textit{\textbf{Clustering Coefficient - Transitivity ($C^\Delta$)}}\\
We calculate the \textit{Clustering Coefficient} ($C^\Delta$) as a measure of the proportion of closed triangles in a graph\cite{newman2003structure}:
\[C^\Delta = 3 \times \frac{\text{number of triangles}}
                    {\text{number of connected triples}}\]

This metric shows us how closely related, or ``clustered,'' that the graph is. It is also the probability that two nodes will be connected if they share a common neighbor\cite{newman2003properties}. In our context, if contributors $C_1$ and $C_2$ both collaborate with a common contributor $C_3$, this metric represents the probability that $C_1$ and $C_2$ will collaborate.

@@

From this \textit{Local Clustering Coefficient} $C^{\lambda}_i$, we can then take the average over the entire set of nodes in the graph.
\[ \overline{C^\lambda} = \frac{1}{|P|}\sum^{|P|}_{i=1}C^{\lambda}_i \]

@@

Where given {$v_i \in P$}, {$v_j \in P$}, the function {$\lambda(v_i,v_j)$} denotes the shortest path between the two nodes. We only considered path length for connected subgraphs, avoiding some complexities of this calculation on disconnected graphs\cite{boccaletti2006complex}.

@@

We assume throughout that any graph exhibiting {$Q > 1$} is a small world graph.

@@

To further study these factors, we arranged interviews with two current professionals in the field. Both have had experience with project management and are certified Project Management Professionals (PMP\textregistered). We proposed to them the following questions, with the intent of expanding further the questions posed above:

\noindent\\\textit{Q1: How do development teams collaborate and share ideas?}\\
\noindent\\\textit{Q2: Do teams generally remain intact over multiple projects, or do members often shuffle and re-combine?}\\
\noindent\\\textit{Q3: When assembling a team, which is preferred -- members familiar with each other, or members familiar with the task or project?}\\
\noindent\\\textit{Q4: What are some characteristics of a high performing team?}\\
\noindent\\\textit{Q5: In your experience, is it better to (A) keep an innovative team intact, (B) periodically bring in new members or re-assign existing members, or (C) split up and disperse the team to as many other projects as possible?}\\

@@

Summary statistics for the graph analysis are captured in tables \ref{fig:freecode_summary_stats} and \ref{fig:sf_summary_stats}.  Statistics analyzed included:\\
$Q$ - \textit{Small World Quotient}\\
$C^\Delta$ - \textit{Clustering Coefficient - Transitivity}\\
$\overline{C^\lambda}$ - \textit{Clustering Coefficient - Average}\\
$L$ - \textit{Average Path Length}\\
$N$ - \textit{Number of Nodes}\\
$E$ - \textit{Number of Edges}\\
The results shown were as measured over the number of projects $P$, with the total number of contributors $C$. The number of connected subgraphs analyzed is shown as $P_c$. The connected subgraphs are the main unit of analysis, and each statistical operation is applied to the set of subgraphs $P_c$.

@@

Distribution of $Q$ values are shown in figures \ref{fig:q_fc_distribution} and \ref{fig:q_sf_distribution}. The values of $Q$ generally follow a power-law distribution, with most values being at our around {$Q = 1$}. This indicates that many graphs are close to the delineation of being considered small world\cite{humphries2008network}.

@@

\subsection{Qualitative}
The research interviews provided a look into how some of these quantitative findings translate into actual industry practice. They also helped us to obtain a deeper understanding of what we had found in the data analysis, as we could begin to build some workplace context into our results. There were several re-occurring themes from which we will base our analysis\cite{stmartin_interview,rana_interview}:

@@

It was also indicated that priority was a major factor. Different project priorities, and the resultant shifting of resources, had the possible effect of weakening a strong team's core members. This can be related to focus as well, as changing priorities can cause resource realignment and decreased overall focus.
